---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# filetree

<!-- badges: start -->
[![R-CMD-check](https://github.com/tjmahr/filetree/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tjmahr/filetree/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

An R package for declarative filetrees with filename validation and parsing.


## Installation

Install the development version of filetree from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("tjmahr/filetree")
```

## Example (filetree with no problems)

Let's say that in our setting, we collect raw data files, process the data and
deposit the files into folders on a shared network drive. We want to make sure
that the directory names and filenames in this directory are consistently named
and correctly organized. With filetree, we describe how these files should be
organized and it checks for any violations to this file organization scheme.

Here is some example longitudinal data organized by participant, time, and 
task. 

```{r demo}
"./inst/demo-1" |> 
  fs::dir_tree(recurse = TRUE)
```

The subject identifiers have a pattern (two letters + hyphen + number), and
these identifiers appear in the filenames in the task folder. Therefore, one
thing filetree should check is that the subject identifiers in the parent folder
and in the filename agree.

We create a filetree with `ft_init()`. Each *layer* is a level of hierarchy. At
the root, when we list the folders, we see subjects. So, the first layer is
`"subject"`. Inside of a subject folder, when we list the folders, we see time
points. The layer is therefore `"time"`. Finally, in the last level, we see our
main task files, so we call this layer `"task"`. 

```{r}
library(filetree)

ft <- ft_init(
  root = "./inst/demo-1", 
  layers = c("subject", "time", "task")
)
ft
```

We define the file organization rules (*schema*) with two ingredients: 

1. field *regexes* define substring values to extract, 
2. *templates* arrange the fields with fixed text.

As a slogan: Regexes define fields, and templates arrange fields. 

Regexes are registered with `ft_add_regex()` by using a named character vector.

```{r}
ft <- ft |>
  ft_add_regex(c(
    subject = "\\w{2}-\\d{2}",
    time = "day\\d{2}",
    task = "red|green"
  )) 
ft
```

Templates are added with `ft_add_dir_template()` or `ft_add_file_template()`. We
can refer to the fields from the regex pool by using `{regex_name}`. Thus, files
at the task layer have the template `"{subject}_{task}.txt"` which then gets
expanded (by filetree) into the regex: `"^(\\w{2}-\\d{2})_(red|green)[.]txt$"`.

```{r}
ft <- ft |>
  ft_add_dir_template(
    layer = "time", 
    template = "{time}"
  ) |> 
  ft_add_dir_template(
    layer = "subject", 
    template = "{subject}"
  ) |> 
  ft_add_file_template(
    layer = "task",  
    template = "{subject}_{task}.txt"
  )
ft
```

Templates match full directory names or filenames, so `"day{time}"` matches
`"day01"` when `time = "\\d{2}"`, but not `"day01b"`. Templates are also literal
or fixed text patterns: `.txt` is a real extension here. In other words, regexes
are not allowed in the templates except through `{}` fields.

We can view the filetree schema as a tree:

```{r}
ft |> ft_schema_tree()
```

Or as a flat list:

```{r}
ft |> ft_schema_flat()
```

We can retrieve the list of filenames in the filetree with `ft_list()`:

```{r}
# Look at the files
ft |> ft_list() |> fs::path_rel()
```

We can parse and validate the file names with `ft_index()`. This function
returns one row per file with columns about the file's layer, template, and
validation status:

```{r}
ft |> ft_index()

ft |> ft_index() |> dplyr::glimpse()
```

All `.`-initial columns were generated by filetree:

  - `.path` - full file path.
  - `.rel` - relative path
  - `.at_layer` - name of the layer containing the file
  - `.filename` - base filename
  - `.layer__subject`, `.layer__time`, `.layer__task` - folder matched at the
    `subject`, `time`, `task` layers.
  - `.file_template` - name of the file template from `ft_add_file_template()`
    that matched this file
  - `.ok` - whether the filepath passed the schema checks (`TRUE`) or whether
    there was a problem (`FALSE`).
  - `.problems` - list column containing diagnostic messages about each file.

The other columns here are the regex fields extracted from each filepath:

  - `subject`
  - `time`
  - `task`


## Example (filetree with problems)

For comparison, here is a file tree with some problems. 

```
# fs::dir_tree("./inst/demo-2")
├── ab-01
│   ├── day01
│   │   ├── ab-01_blue.txt       <---- 'blue' is not a valid task
│   │   └── ab-01_red.txt
│   ├── day02
│   │   ├── ab-01_green.txt
│   │   └── ac-01_red.txt        <---- wrong subject ('ac-01')
│   └── day03
│       └── ab-01_green.txt
└── ac-02
    ├── day01
    │   ├── ac-02_green.txt
    │   └── ac-02_red.txt
    ├── day02
    │   ├── ac-02_green.txt
    │   └── ac-02_red.txt
    └── day3                     <---- expected 2-digit format
        ├── ac-02_green.txt    
        └── ac-02_red.txt
```

That is,

- a misformatted day folder
- a file with the wrong subject
- a file with a nonexisting task ("blue")

We reuse the same templates as earlier. 

```{r}
ft2 <- "./inst/demo-2" |> 
  ft_init(c("subject", "time", "task")) |>
  ft_add_regex(c(
    subject = "\\w{2}-\\d{2}",
    time = "day\\d{2}",
    task = "red|green"
  )) |> 
  ft_add_dir_template("subject", "{subject}") |> 
  ft_add_dir_template("time", "{time}") |> 
  ft_add_file_template("task", "{subject}_{task}.txt")
```

Now, the `.ok` column from `ft_index()` indicates problems:

```{r}
ft2 |> 
  ft_index() |> 
  subset(!.ok)
```

The `.problems` column contains formatted messages about each file, so these 
are best viewed with `ft_glimpse_problems()`:

```{r}
ft2 |> 
  ft_index() |> 
  ft_glimpse_problems()
```


### Template consistency and conditional templates

We are going to add three new complications to the ongoing example.

1.  A new "manifest" file appears at the `time` layer. This is a "sidecar" file.
    It can appear in a layer alongside directories.
2.  `task`-layer files now include `time` values with the template
    `"{subject}_{time}_{task}.txt"`.
3.  There is a `"yellow"` task that only appears on `day03`. 

Each of these new additions is broken somehow in the following example:

```
# fs::dir_tree("./inst/demo-3")
./inst/demo-3
├── ab-01
│   ├── ab-01-manifest.txt
│   ├── day01
│   │   ├── ab-01_01_green.txt
│   │   └── ab-01_01_red.txt
│   ├── day02
│   │   ├── ab-01_01_green.txt   <---- wrong time value
│   │   └── ab-01_01_red.txt     <---- wrong time value
│   └── day03
│       └── ab-01_03_yellow.txt
└── ab-02
    ├── aa-02-manifest.txt       <---- wrong subject value
    ├── day01
    │   ├── ab-02_01_green.txt
    │   └── ab-02_01_red.txt
    ├── day02
    │   ├── ab-02_02_green.txt
    │   ├── ab-02_02_red.txt
    │   └── ab-02_02_yellow.txt  <---- can't go here
    └── day03
        └── ab-02_03_green.txt   <---- can't go here
```

We can make templates apply conditionally by using `when = list(...)` to 
restrict a template to certain field values. We can also conditionally override
field regexes using `with = c(...)`. 

```{r}
ft3 <- "./inst/demo-3" |> 
  ft_init(c("subject", "time", "task")) |>
  ft_add_regex(c(
    subject = "\\w{2}-\\d{2}",
    time = "\\d{2}",
    task = "red|green"
  )) |> 
  ft_add_dir_template("time", "day{time}") |> 
  ft_add_dir_template("subject", "{subject}") |> 
  ft_add_file_template(
    "task", 
    "{subject}_{time}_{task}.txt",
    # limit when this template is used
    when = list(time = c("01", "02")),
  ) |> 
  ft_add_file_template(
    "time", 
    "{subject}-manifest.txt"
  ) |> 
  ft_add_file_template(
    "task", 
    c(day03 = "{subject}_{time}_{task}.txt"),
    # limit when this template is used
    when = c(time = "03"),
    # temporarily override a field regex too
    with = c(task = "yellow")
  )
ft3
```

When using an alternative or additional template, it can be helpful to use a
named character vector like `c(day03 = "{subject}_{time}_{task}.txt")` to
indicate something about the nature of this special case. These template names
are provided in the schema views of the filetree:

```{r}
ft_schema_tree(ft3)

ft_schema_flat(ft3)
```

We should find our five problems:

```{r}
ft3 |> 
  ft_index() |> 
  ft_glimpse_problems()
```


## Additional package usage notes

### `ft_index()` can take a vector of file paths

By default, `ft_index()` searches for files:

```{r}
nrow(ft_index(ft))
```

But if that file-listing is time consuming (as in a network location), we 
can store the filename beforehand and pass it into `ft_index()`:

```{r}
files_to_check <- ft_list(ft)
nrow(ft_index(ft, files_to_check))
```


### Integer layers 

Layer arguments also accept positive integer positions. Layer `1` is the first
configured layer, while layer `0` is the implicit root and cannot be used for
template registration.

```{r}
ft |>
  ft_add_dir_template(
    layer = 2, 
    template = c(extra = "__{time}")
  )
```

I added this convention after having a case where there was a "happy" data path 
organization like  `. / {speaker}` / `{task}` and a misfit path 
`. / _excluded / {speaker} / {task}` so it became easier to refer to the misfit 
files' layer depth using numbers.

### Ignoring files and directory subtrees

Some files are intentionally outside the schema. Ignored files are pruned from
`ft_list()` and `ft_index()` by default. Ignored directory templates prune every
file below a matching directory.

```{r}
files_to_check <- fs::path(
  ft$root,
  c(
    "ab-01/day01/ab-01_red.txt",
    "ab-01/day01/ab-01_notes.txt",
    "ab-01/tmp/not-a-data-file.txt"
  )
)

ft_ignored <- ft |>
  ft_ignore_file_template("task", "{subject}_notes.txt") |>
  ft_ignore_dir_template("time", "tmp")

ft_ignored |> 
  ft_index(files_to_check) |> 
  nrow()

ft_ignored |> 
  ft_index(files_to_check, include_ignored = TRUE) |> 
  nrow()
```

### Strict mode

There is an additional `strict` mode for `ft_index()`. By default, when no 
templates are defined for a file in a layer, those files are accepted.

```{r}
ft4 <- "./inst/demo-1" |> 
  ft_init(
    layers = c("subject", "time", "task")
  )

files_to_check <- fs::path(
  ft4$root,
  c("ab-01/day01/ab-01_red.txt", "ab-01/sidecar.txt")
)

ft4 |> 
  ft_index(files_to_check) |> 
  ft_glimpse_problems()
```

When `strict` is set to `TRUE`, then every file (row) must have a template
associated with it:

```{r}
ft4 |> 
  ft_index(files_to_check, strict = TRUE) |> 
  ft_glimpse_problems()
```

My rationale with strict mode is that every file in the tree must be accounted 
for somehow.

## LLM-assisted development

This package has been developed using ChatGPT/Codex with the
[obra/superpowers](https://github.com/obra/superpowers) and the
[posit-dev/skills](https://github.com/posit-dev/skills) skills installed. 
Agents track their notes in `inst/`. This package is my first time using 
LLM agents. The agents are not supposed to edit README.Rmd, so this document 
provides the final word on package functionality after R package tests and 
checks clear.

## Current impressions

I have been using this tool to validate our speech corpus data on our network 
drives. The `ft_index()` operation is very expensive in this case because 
of the network access and the thousands upon thousands of files at play. I've 
found it better to instead do something like:

```{r}
files_to_check <- fs::dir_ls(ft$root, recurse = TRUE, type = "file")
results <- ft_index(ft, files_to_check)

# Even faster to prefilter the tree if we are just interested in a particular
# set of files
files_to_check <- ft$root |> 
  fs::dir_ls(recurse = TRUE, type = "file", regexp = ".txt$")

results <- ft_index(ft, files_to_check)
```


***

It would be nice to 

- [ ] check inventory/completeness. (Did you notice a missing "red" file in the
  first tree?)
  
- [x] constrain parent folder. (Maybe a "yellow" is given on and only on day 3.)
  Or is that more of a dplyr-layer move for validation?

- [ ] add validation that we can reconstruct `.rel` from the concatenation of 
  each layer?
  
- [x] having a really good and fast way to see the problem files
